---
title: "sosta - Comparison"
author: "Samuel Gunz"
date:  last-modified
format: 
    html:
      toc: true
      code-fold: true
      self-contained: true
      highlight-style: github
      code-line-numbers: true
output-dir: "../docs/"
---

# Setup 
```{r}
#| message: false
#| warning: false
library("SpatialExperiment")
library("sosta")
library("ggplot2")
library("tidyr")
library("dplyr")
library("sf")
library("spdep")
library("ggspavis")
library("reticulate")

theme_set(theme_bw())
```

# Structure simulation

Using `sosta::simulateTissueBlobs` and `sosta::createPointPatternTissue` to create a simple simulation of tissue and cell types within.

```{r}
set.seed(38)
tissueImage <- simulateTissueBlobs(256, 42, 11)
```

```{r}
poly <- binaryImageToSF(tissueImage, xmin = 0, xmax = 256,
                ymin = 0, ymax = 256)

plot(poly)
```

```{r}
pointPattern1 <- createPointPatternTissue(tissueImage, 0.1, 0.1, 0.005, 0.005, 0.005)
```

```{r}
pointPattern1 |> 
    as.data.frame() |> 
    ggplot(aes(x = x, y = y, color = marks)) +
    geom_point(size = 1.5, stroke = 0) +
    coord_equal() +
    scale_color_manual(values = pals::okabe(4)[-1])
```

Create `SpatialExperimentObject`

```{r}
df1 <- as.data.frame(pointPattern1)
df1$imageName <- "image1"


tstSPE <- SpatialExperiment::SpatialExperiment(
    colData = df1,
    spatialCoordsNames = c("x", "y")
)
```

# sosta

Using default parameter estimation.

```{r}
res <- reconstructShapeDensityImage(tstSPE, marks = "marks",
                                    markSelect = "A",
                                    imageId = "image1",
                                    imageCol = "imageName")
plot(res)
```
```{r}
# using chatGPT
library(sf)
library(terra)

# Convert sf polygons to one binary matrix over a user defined window
# polys: sf polygon object
# xmin, xmax, ymin, ymax: window extent
# nrow, ncol: matrix dimensions
sf_to_binary_matrix <- function(polys,
                                xmin, xmax,
                                ymin, ymax,
                                nrow, ncol) {
  
  # build raster template from user extent
  r_template <- rast(
    xmin  = xmin,
    xmax  = xmax,
    ymin  = ymin,
    ymax  = ymax,
    nrows = nrow,
    ncols = ncol
  )
  
  # assign value 1 to all polygons
  polys_val <- polys |>
    mutate(value = 1)
  
  # rasterize polygons to binary raster
  r_bin <- polys_val |>
    vect() |>
    rasterize(
      y          = r_template,
      field      = "value",
      background = 0
    )
  
  # return binary matrix
  as.matrix(r_bin, wide = TRUE)
}

# flip matrix in y direction
flip_y <- function(mat) {
  mat[nrow(mat):1, , drop = FALSE]
}

# rotate matrix 90 degrees clockwise
rotate_90_cw <- function(mat) {
  t(mat[nrow(mat):1, , drop = FALSE])
}


# Jaccard index for two binary matrices
# A, B: matrices with values 0 or 1
jaccard_binary <- function(A, B) {
  # check dimensions
  if (!all(dim(A) == dim(B))) {
    stop("A and B must have the same dimensions")
  }
  
  # convert to logical 1 = TRUE
  A1 <- A == 1
  B1 <- B == 1
  
  # intersection size
  inter <- sum(A1 & B1)
  
  # union size
  uni <- sum(A1 | B1)
  
  if (uni == 0) {
    return(NA_real_)  # undefined if both empty
  }
  
  return(inter / uni)
}
```



```{r}
res_sosta <- sf_to_binary_matrix(res, 0, 256, 0, 256, 256, 256) |> rotate_90_cw()
jaccard_binary(res_sosta, tissueImage)
```


# imcRtools

Using `patchDetection` and corresponding `patchSize` to get the polygon.
```{r}
library("imcRtools")

tstSPE_icr <- buildSpatialGraph(tstSPE,
                                coords = c("x", "y"),
                                img_id = "imageName",
                                type = "expansion",
                                threshold = 10)

tstSPE_icr <- patchDetection(tstSPE_icr, 
                             coords = c("x", "y"),
                             patch_cells = tstSPE$marks == "A",
                             colPairName = "expansion_interaction_graph",
                             expand_by = 5, 
                             img_id = "imageName")
 
plotSpatial(tstSPE_icr, 
            coords = c("x", "y"),
            img_id = "imageName", 
            node_color_by = "patch_id",
            scales = "free")
```


```{r}
p <- patchSize(tstSPE_icr,
               coords = c("x", "y"))

res_imcr <- st_as_sfc(p$polygon) |> st_as_sf()
plot(res_imcr)

res_imcr_b <- sf_to_binary_matrix(res_imcr, 0, 256, 0, 256, 256, 256) |>
  rotate_90_cw()
```


```{r}
jaccard_binary(res_imcr_b, tissueImage)
```

# SPIAT

```{r}
library("SPIAT")
# See Issue: https://github.com/TrigosTeam/SPIAT/issues/45
colnames(spatialCoords(tstSPE)) <- c("Cell.X.Position", "Cell.Y.Position")

spe_border <- identify_bordering_cells(
    tstSPE,
    reference_cell = "A",
    feature_colname = "marks"
)

spe_dist <- calculate_distance_to_margin(spe_border)

spe_structure <- define_structure(
    spe_dist,
    cell_types_of_interest = c("A", "B", "C"),
    feature_colname = "Cell.Type",
    n_margin_layers = 5
)

plot_cell_categories(spe_structure, feature_colname = "Structure")
```

Manually change code to extract a polygon
```{r}
library("alphahull")
# adapted from https://github.com/TrigosTeam/SPIAT/blob/devel/R/identify_bordering_cells.R
identify_bordering_cells_poly <- function(spe_object, reference_cell, 
                                     feature_colname = "Cell.Type",
                                     ahull_alpha = NULL,n_to_exclude = 10, 
                                     plot_final_border = TRUE){
    # CHECK
    if (is.null(SummarizedExperiment::colData(spe_object)[,feature_colname])){
        stop("Undefined column name. Please check if input contains `feature_colname`!")
    }
    if (!(reference_cell %in% 
          SummarizedExperiment::colData(spe_object)[,feature_colname])){
        stop("Reference cell not found!")
    }
    
    coords_df <- data.frame(SpatialExperiment::spatialCoords(spe_object))
    
    ## draw boundaries
    l <- list()
    draw.polys <- data.frame("x" = c(max(coords_df$Cell.X.Position), 
                                     max(coords_df$Cell.X.Position),
                                     min(coords_df$Cell.X.Position), 
                                     min(coords_df$Cell.X.Position)),
                             "y" = c(min(coords_df$Cell.Y.Position), 
                                     max(coords_df$Cell.Y.Position),
                                     max(coords_df$Cell.Y.Position), 
                                     min(coords_df$Cell.Y.Position)))
    poly <- sp::Polygon(draw.polys, hole = FALSE)
    l[[1]] <- poly
    
    polys <- sp::Polygons(l,ID = c("a"))
    sp_obj <- sp::SpatialPolygons(list(polys))
    
    # for loop, get the boundary cells and inside cells for each polygon #####
    data <- data.frame(SummarizedExperiment::colData(spe_object))
    data <- cbind(data,data.frame(SpatialExperiment::spatialCoords(spe_object)))
    data[,"Region"] <- "Outside"
    
    # get the coords
    buffered_polygon <- methods::slot(sp_obj@polygons[[1]]@Polygons[[1]],
                                      "coords")
    
    # identify the cells in the drawn polygon
    inpolygon <- sp::point.in.polygon(
        data$Cell.X.Position, data$Cell.Y.Position, buffered_polygon[, 1], 
        buffered_polygon[, 2])
    allcells_in_polygon <- data[which(inpolygon!= 0),
                               c(feature_colname,"Cell.X.Position",
                                 "Cell.Y.Position",feature_colname)]
    tumour_in_polygon <- 
        allcells_in_polygon[which(allcells_in_polygon[,feature_colname] == 
                                      reference_cell),]
    tumour_in_polygon <- unique(tumour_in_polygon)
    
    # ahull of the tumour cells
    # define the value of alpha
    requireNamespace("alphahull", quietly = TRUE)
    if (is.null(ahull_alpha)){
        n_cells <- dim(tumour_in_polygon)[1]
        if (n_cells<200){
            alpha <- 60} else if (n_cells > 5000){
            alpha <- 90} else {
            alpha <- (n_cells - 300)/160 + 60}
        methods::show(paste("The alpha of Polygon is:", alpha))
        ahull <- alphahull::ahull(tumour_in_polygon$Cell.X.Position,
                                 tumour_in_polygon$Cell.Y.Position, 
                                 alpha = alpha)
    }
    else {
        ahull <- alphahull::ahull(tumour_in_polygon$Cell.X.Position,
                                 tumour_in_polygon$Cell.Y.Position, 
                                 alpha = ahull_alpha)
    }
    # fix ahull
    resahull <- ahull
    ahull <- fix_ahull(ahull)
    # my polygon function
    xahull <- ahull$xahull
    arc <- ahull$arcs
    ahull_polygon <- get_polygon(xahull,arc,n_to_exclude)
    
    # identify the cells that compose the ahull
    border_ids <- c()
    for (i in c(seq_len(length(ahull_polygon)))){
        p <- data.frame(ahull_polygon[[i]])
        colnames(p) <- c("Cell.X.Position","Cell.Y.Position")
        
        # find the bordering cells in the original dataset to find the IDs
        common_cells <- dplyr::intersect(
            data[,c("Cell.X.Position","Cell.Y.Position")],
            p[,c("Cell.X.Position","Cell.Y.Position")])
        border_ids <- c(border_ids, rownames(common_cells))}
    
    # identify the cells that are in the ahull
    points_in_polygon <- data.frame()
    for (i in c(seq_len(length(ahull_polygon)))){
        p <- ahull_polygon[[i]]
        in_p <- sp::point.in.polygon(allcells_in_polygon$Cell.X.Position, 
                                     allcells_in_polygon$Cell.Y.Position, 
                                     p[,1], p[,2])
        points_in_polygon <- 
            unique(rbind(
                points_in_polygon, 
                allcells_in_polygon[which(in_p == 1),
                                    c(feature_colname, "Cell.X.Position",
                                      "Cell.Y.Position",
                                      feature_colname)]))}
    points_in_polygon_df <- as.data.frame(points_in_polygon)
    cells_in_boundary <- points_in_polygon_df
    
    in_border_ids <- rownames(cells_in_boundary)
    data[in_border_ids,"Region"] <- "Inside"
    data[border_ids,"Region"] <- "Border"

    # plot and return #####
    SummarizedExperiment::colData(spe_object)$Region <- data[,"Region"]
    # check if bordering cells are detected and if plot the border
    if (dim(data[which(data$Region=="Border"),])[1]!=0){
        if (isTRUE(plot_final_border)){
            plot(data[which(data$Region=="Border"), 
                      c("Cell.X.Position","Cell.Y.Position")], 
                 pch = 19, cex = 0.3, main = paste(attr(spe_object, "name"),
                                                   "bordering cells"))}
    }
    else{
        warning("There are no bordering cells detected when alpha = ", ahull_alpha)
        }
    # save the number of clusters identified
    attr(spe_object, "n_of_clusters") <- length(ahull_polygon)
    return(c(spe = spe_object, ahull = resahull))
}

# from: https://github.com/TrigosTeam/SPIAT/blob/d5dedad7fe4ae645940f3c80e6343347360b1618/R/utils.R#L65
# fix the bug in `ahull` function from `alphahull` package
fix_ahull <- function(ahull){ # the order of cells returned by ahull is messy
    # this function reorders the cells
    arc <- ahull$arcs
    n_cells <- dim(arc)[1]
    # copy the arc ends
    ends <- arc
    ends <- rbind(ends, ends[1,])
    i <- 1
    while (i < n_cells){
        end2 <- ends[i,8]
        next_end1 <- ends[i+1,7]
        next_end2 <- ends[i+1,8]
        while (end2 != next_end1){ # the connection breaks here
            if (next_end2 == next_end1) {
                # if the next cell is a loner cell, or the current cell is the last cell,
                # the current cell does not have to be issue cell
                i <- i+1
                if (i >= (dim(arc)[1])) break
                next_end1 <- arc[i+1,7]
                next_end2 <- arc[i+1,8]
            }
            
            else{ # the next cell is not loner cell, confirm the current cell is the issue cell
                # break the current loop, go to next cell (skip loner cells)
                for (j in (i+1):n_cells){
                    t_end1 <- ends[j,7]
                    t_end2 <- ends[j,8]
                    # found the correct cell!
                    if (t_end1 == end2 || t_end2 == end2){ 
                        ends[n_cells+1,] <- ends[i+1,]
                        ends[i+1,] <- ends[j,]
                        ends[j,] <- ends[n_cells+1,]
                    }
                    if (t_end2 == end2){
                        ends[i+1,8] <- t_end1
                        ends[i+1,7] <- end2
                    }
                }
                break
            }
        }
        # next cell
        i <- i+1
    }
    ends <- ends[-(n_cells+1),]
    ahull$arcs <- ends
    return(ahull)
}

# gets the coordinates of points on ahull
get_polygon <- function(xahull, arc, n_to_exclude){ 
    df_list <- list()
    n <- 0
    s <- 1
    for (i in seq_len(dim(arc)[1]-1)){
        if (arc[i,8] != arc[i+1,7]){
            if (i-s > 4){
                n <- n+1
                df_list[[n]] <- arc[c(s:i),]
            }
            s <- i+1
        }
    }
    if (length(df_list) == 0) df_list[[1]] <- arc
    poly_list <- list()
    c <- 0 
    for (j in c(seq_len(length(df_list)))){
        df <- df_list[[j]]
        # eliminate the small tumour clusters (add later)
        if (dim(df)[1] <= n_to_exclude){
            next
        }
        else{
            c <- c+1
            cell_ID <- c()
            locs <- c()
            for (i in seq_len(dim(df)[1])){
                cell_ID <- df[i,7]
                locs <- rbind(locs,xahull[cell_ID,c(1,2)])
            }
            poly_list[[c]] <- locs
        }
    }
    return(poly_list)
}
```

Extract and plot polygon

```{r}
res_spiat <- identify_bordering_cells_poly(
    tstSPE,
    reference_cell = "A",
    feature_colname = "marks",
    plot_final_border = FALSE
)

plot(res_spiat$ahull.ashape.obj,
     asp = 1,
     col = c('darkblue', 'grey75'))
```
```{r}
ashape_to_polygon <- function(ashape){
  # Get the data, edges of the connected lines
  data <-  ashape$edges[,c('x1','y1','x2','y2')]
  # Create list of edges
  result_list <- lapply(1:nrow(data), function(i) {
    t(matrix(data[i,], ncol = 2))
  })
  # Create multilinestring that connects the edges
  mls <- st_multilinestring(result_list)
  poly <- st_polygonize(mls)
  # Extract polygons from st collecion
  ashape_poly <- st_collection_extract(poly, "POLYGON")
  return(st_as_sf(st_sfc(ashape_poly)))
}


spiat_poly <- ashape_to_polygon(res_spiat$ahull.ashape.obj)
plot(spiat_poly)
```
```{r}
res_bin_spiat <- sf_to_binary_matrix(spiat_poly, 0, 256, 0, 256, 256, 256) |>
  rotate_90_cw()

jaccard_binary(res_bin_spiat, tissueImage)
```

# Cell Charter

Using default values 
```{python}
# created by chatGPT
import numpy as np
import pandas as pd
import anndata as ad
import squidpy as sq
import cellcharter as cc
import matplotlib.pyplot as plt

def table_to_adata(df: pd.DataFrame) -> ad.AnnData:
    df = df.copy()

    # one row per cell
    df.index = df.index.astype(str)

    # minimal X so AnnData is valid
    X = np.ones((df.shape[0], 1), dtype=np.float32)

    adata = ad.AnnData(X=X)

    # cell labels and image id
    adata.obs["marks"] = df["marks"].astype("category")
    adata.obs["imageName"] = df["imageName"].astype("category")

    # spatial coordinates
    adata.obsm["spatial"] = df[["x", "y"]].to_numpy(dtype=np.float32)
    return adata


# build adata
adata = table_to_adata(r["df1"])

adata
```

```{python}
sq.gr.spatial_neighbors(adata, library_key='imageName', coord_type='generic', delaunay=True)
```


```{python}
sq.pl.spatial_scatter(
    adata,
    color='marks', 
    connectivity_key="spatial_connectivities",
    library_id=['test'],
    shape=None
)
plt.show()
```

```{python}
adata_A = adata[adata.obs["marks"] == "A"].copy()

cc.gr.connected_components(adata_A, cluster_key='marks')
cc.tl.boundaries(adata_A, min_hole_area_ratio=0.01)
```

```{python}
# created by chatGPT
import matplotlib.pyplot as plt
import numpy as np

boundaries = adata_A.uns["shape_component"]["boundary"]
coords = adata_A.obsm["spatial"]

# categorical marks
marks = adata_A.obs["marks"].astype("category")
codes = marks.cat.codes
categories = marks.cat.categories

fig, ax = plt.subplots()

# scatter colored by mark
sc = ax.scatter(
    coords[:, 0],
    coords[:, 1],
    c=codes,
    s=5
)

# plot component boundaries
for comp_id, poly in boundaries.items():
    x, y = poly.exterior.xy
    ax.plot(x, y, linewidth=2)

ax.set_aspect("equal")
plt.show()
```

## Extract Python polygons

```{python}
from shapely.geometry import MultiPolygon, Polygon

# collect all Polygon geometries
polys = [v for v in boundaries.values() if isinstance(v, Polygon)]

# build one MultiPolygon
boundaries_mpoly = MultiPolygon(polys)
type(boundaries_mpoly)
```


```{r}
p0 <- py$boundaries_mpoly 
print(p0)


p0_poly <- st_as_sfc(p0$wkt) |> st_as_sf()
plot(p0_poly)
```

```{r}
sf_cell_charter <- sf_to_binary_matrix(p0_poly, 0, 256, 0, 256, 256, 256) |>
  rotate_90_cw()

jaccard_binary(sf_cell_charter, tissueImage)
```

# GRIDGENE

```{python}
# created by chatGPT
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

from gridgene import get_arrays as ga
from gridgene import contours as gg_contours
from gridgene import get_masks as gg_masks


def _plot_contour(ax, contour):
    # shapely polygon
    if hasattr(contour, "exterior"):
        x, y = contour.exterior.xy
        ax.plot(x, y, linewidth=2)
        return

    # numpy coordinates from gridgene, usually (row, col)
    c = np.asarray(contour)
    if c.ndim == 3 and c.shape[-2] == 1 and c.shape[-1] == 2:
        c = c[:, 0, :]
    if c.ndim != 2 or c.shape[1] != 2:
        return

    row = c[:, 0]
    col = c[:, 1]
    ax.plot(col, row, linewidth=2)  # plot as (x=col, y=row)


def gridgene_segment_marks(
    df: pd.DataFrame,
    library_id: str = "image1",
    mark_tum: str = "A",
    kernel_size: int = 80,
    density_th: int = 40,
    min_area: int = 1000,
    plot: bool = True,
):
    d = df[df["imageName"] == library_id].copy()

    # GRIDGEN expects integer pixel coords in X, Y and labels in target
    d["X"] = np.rint(d["x"]).astype(int)
    d["Y"] = np.rint(d["y"]).astype(int)
    d["target"] = d["marks"].astype(str)

    # image arrays are (rows=Y, cols=X)
    height = int(d["Y"].max() + 1)
    width = int(d["X"].max() + 1)

    target_dict = {t: i for i, t in enumerate(sorted(d["target"].unique()))}

    array_total = ga.transform_df_to_array(
        df=d,
        target_dict=target_dict,
        array_shape=(height, width, len(target_dict)),
    ).astype(np.int8)

    df_tum, array_tum, _ = ga.get_subset_arrays(
        d,
        array_total,
        target_dict,
        target_list=[mark_tum],
        target_col="target",
    )

    CTum = gg_contours.ConvolutionContours(array_tum, contour_name=f"mark_{mark_tum}")
    CTum.get_conv_sum(kernel_size=kernel_size, kernel_shape="square")
    CTum.contours_from_sum(
        density_threshold=density_th,
        min_area_threshold=min_area,
        directionality="higher",
    )

    CAll = gg_contours.ConvolutionContours(array_total, contour_name="all_marks")
    CAll.get_conv_sum(kernel_size=kernel_size, kernel_shape="square")
    CAll.contours_from_sum(
        density_threshold=density_th,
        min_area_threshold=min_area,
        directionality="higher",
    )

    GM = gg_masks.GetMasks(image_shape=(height, width))
    mask_all = GM.create_mask(CAll.contours)
    mask_tum = GM.create_mask(CTum.contours)
    mask_tum = GM.fill_holes(mask_tum)

    if plot:
        fig, axs = plt.subplots(1, 2, figsize=(12, 5))

        # left. points colored by marks + contours
        marks = d["target"].astype("category")
        codes = marks.cat.codes
        cats = marks.cat.categories

        axs[0].scatter(
            d["X"].to_numpy(),
            d["Y"].to_numpy(),
            c=codes,
            s=3,
            alpha=0.6,
        )

        for cat in cats:
            axs[0].scatter([], [], label=cat)
        axs[0].legend(title="marks")

        for c in CTum.contours:
            _plot_contour(axs[0], c)

        axs[0].set_aspect("equal")
        axs[0].set_title(f"Contours for mark {mark_tum}")

        # right. masks in same coordinate system as scatter
        extent = [0, width, 0, height]
        axs[1].imshow(mask_all, alpha=0.5, origin="lower", extent=extent)
        axs[1].imshow(mask_tum, alpha=0.5, origin="lower", extent=extent)
        axs[1].set_aspect("equal")
        axs[1].axis("off")
        axs[1].set_title("Mask all and mask mark")

        plt.show()

    return {
        "CTum": CTum,
        "CAll": CAll,
        "mask_tum": mask_tum,
        "mask_all": mask_all,
        "shape": (height, width),
        "target_dict": target_dict,
        "array_total": array_total,
        "array_tum": array_tum,
        "df_image": d,
        "df_tum": df_tum,
    }
```

Needs manual tuning of parameters!!

```{python}
res = gridgene_segment_marks(
    r["df1"],
    library_id="image1",
    mark_tum="A",
    kernel_size=20,    
    density_th=20,      
    min_area=100, 
    plot=True
)
```

```{python}
from shapely.geometry import Polygon, MultiPolygon
import numpy as np

def contours_to_multipolygon(contours):
    polys = []

    for c in contours:
        c = np.asarray(c)

        # reshape if (N,1,2)
        if c.ndim == 3 and c.shape[-2] == 1 and c.shape[-1] == 2:
            c = c[:, 0, :]

        if c.ndim != 2 or c.shape[1] != 2:
            continue

        # convert (row, col) -> (x=col, y=row)
        height = res["shape"][0]
        xy = np.column_stack([c[:,1], height - c[:,0]])

        if len(xy) >= 3:
            poly = Polygon(xy)
            if poly.is_valid:
                polys.append(poly)

    return MultiPolygon(polys)


# tumor polygons
mpoly_tum = contours_to_multipolygon(res["CTum"].contours)

# all polygons
mpoly_all = contours_to_multipolygon(res["CAll"].contours)

type(mpoly_tum)
```

```{python}
from shapely.geometry import Polygon, MultiPolygon
import numpy as np

def contours_to_polygons(contours, height=None, flip_y=False,
                         fix_invalid=True, min_vertices=3, min_area=0.0):
    polys = []
    skipped = []

    for i, raw in enumerate(contours):
        arr = np.asarray(raw)

        # unify to a list of 2D arrays, each (N, 2)
        parts = []

        if arr.ndim == 2 and arr.shape[1] == 2:
            parts = [arr]
        elif arr.ndim == 3:
            # shapes like (M, N, 2) or (M, N, 1, 2) flattened by caller
            for p in arr:
                p = np.asarray(p)
                if p.ndim == 2 and p.shape[1] == 2:
                    parts.append(p)
                elif p.ndim == 3 and p.shape[-2] == 1 and p.shape[-1] == 2:
                    parts.append(p[:, 0, :])
        elif arr.ndim == 4 and arr.shape[-2] == 1 and arr.shape[-1] == 2:
            # shapes like (M, N, 1, 2)
            for p in arr:
                parts.append(p[:, 0, :])

        if not parts:
            skipped.append((i, "no usable part"))
            continue

        for p in parts:
            if p.shape[0] < min_vertices:
                skipped.append((i, "too few vertices"))
                continue

            # row, col -> x, y
            row = p[:, 0]
            col = p[:, 1]

            if flip_y and height is not None:
                y = height - row
            else:
                y = row

            xy = np.column_stack([col, y])

            poly = Polygon(xy)

            if not poly.is_valid and fix_invalid:
                poly = poly.buffer(0)

            if not poly.is_valid:
                skipped.append((i, "invalid"))
                continue

            if poly.area < min_area:
                skipped.append((i, "too small"))
                continue

            polys.append(poly)

    return polys, skipped
```

```{python}
height = res["shape"][0]

polys_tum, skipped_tum = contours_to_polygons(
    res["CTum"].contours,
    height=height,
    flip_y=False,      # or True if you need flip
    fix_invalid=True,
    min_vertices=3,
    min_area=0.0
)

mpoly_tum = polys_tum

print("total contours:", len(res["CTum"].contours))
print("polygons kept:", len(polys_tum))
print("skipped (first 10):", skipped_tum[:10])
```

```{python}
boundaries_mpoly = mpoly_tum
```


```{r}
library(sf)

# extract WKT from each Python geometry
wkt_vec <- vapply(py$boundaries_mpoly, function(g) g$wkt, character(1))

# create sfc
sfc_grid <- st_as_sfc(wkt_vec, crs = NA)

# optional. convert to sf
sf_grid <- st_as_sf(data.frame(id = seq_along(sfc_grid)),
                     geometry = sfc_grid)

plot(sf_grid)
```


```{r}
res_gridgene_b <- sf_to_binary_matrix(sf_grid, 0, 256, 0, 256, 256, 256) |>
     rotate_90_cw()
jaccard_binary(res_gridgene_b, tissueImage)
```





