---
title: "Reconstruction and quantification of pancreatic islets from IMC data"
author: "Samuel Gunz"
date:  last-modified
format: 
    html:
      toc: true
      self-contained: true
      highlight-style: github
      code-line-numbers: true
---


```{r}
#| message: false
#| warning: false

library("sosta")
library("ggplot2")
library("dplyr")
library("tidyr")
library("SpatialExperiment")
library("ExperimentHub")
library("lme4")
library("lmerTest")
library("ggfortify")
library("dplyr")


set.seed(8048)
```


Note: part of this code overlaps the `r BiocStyle::Biocpkg('sosta')` vignette "2. Reconstruction and analysis of pancreatic islets from IMC data" and the OSTA chapter "Structure-based analysis" (https://lmweber.org/OSTA/pages/mult-structure-based.html).

### Dataset

```{r, load-data-structs}
#| message: false
# load experiment hub
eh <- ExperimentHub()
oid <- names(eh[eh$title == "Damond_2019_Pancreas - sce - v1 - full"])


# Load single cell experiment object
spe <- eh[[oid]]
# Convert to spatial experiment object
spe <- toSpatialExperiment(spe,
                           sample_id = "image_name",
                           spatialCoordsNames = c("cell_x", "cell_y"))
# Remove as not needed
assays(spe) <- list()
```


```{r, plot-data}
selPl <- sample(unique(spe$image_name), 4)
df <- cbind(colData(spe[, spe$image_name %in% selPl]),
            spatialCoords(spe[, spe$image_name %in% selPl]))
df |>
  as.data.frame() |>
  ggplot(aes(x = cell_x, y = cell_y, color = cell_category)) +
  geom_point(size = 0.5) +
  facet_wrap( ~ image_name, ncol = 2) +
  coord_equal() +
  theme_classic() +
  scale_color_manual(values=unname(pals::cols25(n = 5)))
```

### Islet reconstrution

```{r, estimate-params}
n <- estimateReconstructionParametersSPE(
    spe,
    marks = "cell_category",
    imageCol = "image_name",
    markSelect = "islet",
    nImages = 25,
    nCores = 8
)

thresSPE <- mean(n$thres)
bndwSPE <- mean(n$bndw)
```

```{r}
start <- Sys.time()
# Run on all images
allIslets <- reconstructShapeDensitySPE(
    spe,
    marks = "cell_category",
    imageCol = "image_name",
    markSelect = "islet",
    bndw = bndwSPE,
    thres = thresSPE,
    nCores = 8
)
print(Sys.time() - start)
```

Let's inspect the reconstruction in the sample images.
 
```{r}
df <- cbind(colData(spe[, spe$image_name %in% selPl]),
            spatialCoords(spe[, spe$image_name %in% selPl]))
df |>
  as.data.frame() |>
  ggplot(aes(x = cell_x, y = cell_y, color = cell_category)) +
  geom_point(size = 0.25) +
  facet_wrap(~ image_name, ncol = 2) +
  coord_equal() +
  theme_classic() +
  scale_color_manual(values=unname(pals::cols25(n = 5))) +
  geom_sf(
    data = allIslets[allIslets$image_name %in% selPl, ],
    inherit.aes = FALSE,
    color = "black",
    linetype = "11",
    fill = NA,
    linewidth = 0.75
  ) 
```


```{r}
colsKeep <- c(
  "patient_stage",
  "tissue_slide",
  "tissue_region",
  "patient_id",
  "patient_disease_duration",
  "patient_age",
  "patient_gender",
  "sample_id"
)

# specify factor levels
lv <- c("Non-diabetic", "Onset", "Long-duration")

patientData <- colData(spe) |>
  as_tibble() |>
  group_by(image_name) |>
  select(all_of(colsKeep)) |>
  mutate(patient_stage = factor(patient_stage, levels = lv)) |>
  unique()

allIslets <- allIslets |>
  dplyr::left_join(patientData, by = "image_name")
```


### Quantification of geometric features

```{r}
isletMetrics <- totalShapeMetrics(allIslets)
allIslets <- allIslets |>
  cbind(t(isletMetrics))
```

```{r}
autoplot(
  prcomp(t(isletMetrics), scale. = TRUE),
  x = 1,
  y = 2,
  data = allIslets,
  color = "patient_stage",
  size = 2,
  # shape = 'type',
  loadings = TRUE,
  loadings.colour = "steelblue3",
  loadings.label = TRUE,
  loadings.label.size = 3,
  loadings.label.repel = TRUE,
  loadings.label.colour = "black"
) +
  theme_bw() +
  coord_fixed() +
  scale_color_manual(values = unname(pals::tol(n = 3)))
```
```{r, fig.width=8, fig.height=6}
allIslets |>
  sf::st_drop_geometry() |>
  select(patient_stage, rownames(isletMetrics)) |>
  pivot_longer(-patient_stage) |>
  filter(name %in% c("Area", "Compactness", "Curl")) |>
  ggplot(aes(x = patient_stage, y = value, fill = patient_stage)) +
  geom_violin() +
  geom_boxplot(aes(fill = NULL), width = 0.3) +
  facet_wrap( ~ name, scales = "free") +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  guides(fill = "none") +
  theme_bw() +
  scale_fill_manual(values = unname(pals::tol(n = 3)))
```
```{r, fig.width=8, fig.height=6}
allIslets |>
  sf::st_drop_geometry() |>
  select(patient_stage, patient_id,  rownames(isletMetrics)) |>
  pivot_longer(-c(patient_stage, patient_id)) |>
  filter(name %in% c("Area")) |>
  ggplot(aes(x = as.factor(patient_id), y = value, fill = patient_stage)) +
  geom_violin() +
  geom_boxplot(aes(fill = NULL), width = 0.3) +
  facet_wrap( ~ patient_stage, scales = "free") +
  scale_y_log10() +
  #scale_x_discrete(guide = guide_axis(n.dodge = 2)) +
  guides(fill = "none") +
  theme_bw() +
  scale_fill_manual(values = unname(pals::tol(n = 3)))
```

## Quantification of structure level features

```{r}
assign <- assingCellsToStructures(spe, allIslets,
    imageCol = "image_name", nCores = 8
)
spe$structAssign <- assign
```

```{r}
props <- cellTypeProportions(spe, "structAssign", "cell_type")

colnames(props) <- paste0("prop_", colnames(props))
props$structID <- rownames(props)

allIslets <- left_join(allIslets, props, by = "structID")
```


```{r}
spe$minDist <- minBoundaryDistances(
    spe = spe, imageCol = "image_name",
    structColumn = "structAssign", allStructs = allIslets,
    nCores = 8
)
```
```{r}
library("parallel")
library("dplyr")
library("sf")

minCellTypeStructDist <- function(spe, allStructs, structID = "structID",
                                        cellTypeColumn, imageCol, nCores = 1) {
    # Input checks
    stopifnot(
        "'spe' must be a SpatialExperiment object" = inherits(spe, "SpatialExperiment"),
        "'allStructs' must be an sf object" = inherits(allStructs, "sf"),
        "'structID' must be a single character string and in allStructs" =
            is.character(structID) && length(structID) == 1 && structID %in% colnames(allStructs),
        "'cellTypeColumn' must exist in colData(spe)" =
            cellTypeColumn %in% colnames(colData(spe)),
        "'imageCol' must exist in colData(spe) and allStructs" =
            imageCol %in% colnames(colData(spe)) && imageCol %in% colnames(allStructs)
    )

    # Extract unique image names and remove NAs
    images <- unique(spe[[imageCol]])
    images <- images[!is.na(images)]

    # Convert spe to data frame with spatial coords + metadata
    df <- .SPE2df(spe, imageCol, marks = cellTypeColumn)

    # Split by image
    ls <- split(df, as.factor(df[[imageCol]]))

    # Parallel processing
    res <- mclapply(ls, function(dfSel) {
        selImage <- unique(dfSel[[imageCol]])
        subStruct <- allStructs[allStructs[[imageCol]] %in% selImage, ]

        if (nrow(subStruct) == 0) {
            return(NULL)
        }

        spatialCoordsSf <- st_as_sf(dfSel[, c(1, 2)],
                                    coords = c(colnames(dfSel)[1], colnames(dfSel)[2])
        )

        dist <- st_distance(spatialCoordsSf, st_boundary(subStruct))
        colnames(dist) <- st_drop_geometry(subStruct)[, structID]

        # dfSel[, 4] is assumed to be cellTypeColumn
        resDist <- aggregate(dist ~ dfSel[, 4], FUN = min)
        colnames(resDist)[1] <- cellTypeColumn

        return(resDist)
    }, mc.cores = nCores)

    # Combine to one df, NA when cell type no present in one sample
    resDf <- Reduce(function(x, y) full_join(x, y, by = cellTypeColumn), res)
    # Change orientation of resDf
    rownames(resDf) <- resDf[[cellTypeColumn]]
    return(as.data.frame(t(resDf[,-1])))
}
```

```{r}
mdstr <- minCellTypeStructDist(spe = spe, imageCol = "image_name",
                               cellTypeColumn =  "cell_type",
                               allStructs = allIslets, nCores = 4)

colnames(mdstr) <- paste0("dist_", colnames(mdstr))
mdstr$structID <- rownames(mdstr)

allIslets <- left_join(allIslets, mdstr, by = "structID")
```



```{r}
library(vegan)

structColumn <- "structAssign"
cellTypeColumn <- "cell_type"
nCores <- 3


df <- colData(spe)[, c(structColumn, cellTypeColumn)]
df <- df[!is.na(df[[structColumn]]), ]

# Split data frame
ls <- split(df, as.factor(df[, 1]))

# Compute the proportion of each cell type within each structure
res <- mclapply(ls, function(dfSub) {
    # Compute the frequency and normalize
    idx <- diversity(table(dfSub[[cellTypeColumn]]), index = "shannon")
    return(idx)
}, mc.cores = nCores)

# Combine into single df and name with structs id
resComb <- do.call(rbind, res) |> as.data.frame()

colnames(resComb) <- "shannonDiv"
```

```{r}
resComb$structID <- rownames(resComb)

allIslets <- left_join(allIslets, resComb, by = "structID")
```

# Boders

```{r}
borders <- mclapply(
    st_geometry(allIslets),
    \(x) st_difference(st_buffer(x, 15), st_buffer(x, -15)),
    mc.cores = nCores
) |>
    st_as_sfc() |>
    st_as_sf() # both functions needed for proper conversion

borders$image_name <- allIslets$image_name

```

```{r}
df <- cbind(colData(spe[, spe$image_name %in% selPl]),
            spatialCoords(spe[, spe$image_name %in% selPl]))
df |>
  as.data.frame() |>
  ggplot(aes(x = cell_x, y = cell_y, color = cell_category)) +
  geom_point(size = 0.25) +
  facet_wrap(~ image_name, ncol = 2) +
  coord_equal() +
  theme_classic() +
  scale_color_manual(values=unname(pals::cols25(n = 5))) +
  geom_sf(
    data = borders[borders$image_name %in% selPl, ],
    inherit.aes = FALSE,
    color = "gray25",
    fill = NA,
    linewidth = 0.75,
    linetype = "11"
  ) 
```


```{r}
borders$image_name <- allIslets$image_name
borders$borderID <- paste0("border_", allIslets$structID)
allIslets$borderID <- borders$borderID


spe$IsletBorder <- assingCellsToStructures(spe,
    borders,
    imageCol = "image_name",
    uniqueId = "borderID",
    nCores = 4
)
```

```{r}
structColumn <- "IsletBorder"
cellTypeColumn <- "cell_type"
nCores <- 3


df <- colData(spe)[, c(structColumn, cellTypeColumn)]
df <- df[!is.na(df[[structColumn]]), ]

# Split data frame
ls <- split(df, as.factor(df[, 1]))

# Compute the proportion of each cell type within each structure
res <- mclapply(ls, function(dfSub) {
    # Compute the frequency and normalize
    idx <- diversity(table(dfSub[[cellTypeColumn]]), index = "shannon")
    return(idx)
}, mc.cores = nCores)

# Combine into single df and name with structs id
resComb <- do.call(rbind, res) |> as.data.frame()

colnames(resComb) <- "shannonDivBorder"
resComb$borderID <- rownames(resComb)

allIslets <- left_join(allIslets, resComb, by = "borderID")

```


# Save spe object


```{r}
saveRDS(allIslets, "../../data/allIslets.rds" )
saveRDS(spe, "../../data/spe_islets_sosta.rds" )
```


```{r}
sessionInfo()
```

<!-- ### Between sample comparison using mixed effects models -->

<!-- ## 4th root transform -->

<!-- ```{r} -->
<!-- hist(allIslets$Area^(1/4)) -->
<!-- mod <- lmer(Area^(1/4) ~ patient_stage + (1 | patient_id) + (1 | image_name), data = allIslets) -->
<!-- ``` -->

<!-- #### Model diagnositcs -->

<!-- ::: {.panel-tabset} -->

<!-- ## Residuals vs. fitted  -->

<!-- ```{r} -->
<!-- plot( -->
<!--   mod, -->
<!--   resid(., scaled = TRUE) ~ fitted(.), -->
<!--   col = allIslets$patient_id, -->
<!--   pch = 12, -->
<!--   abline = 0, -->
<!--   xlab = "Fitted values", -->
<!--   ylab = "Standardised residuals" -->
<!-- ) -->
<!-- ``` -->

<!-- ## Residuals vs. fitted per sample -->

<!-- ```{r} -->
<!-- plot( -->
<!--   mod, -->
<!--   resid(., scaled = TRUE) ~ fitted(.) | patient_id, -->
<!--   abline = 0, -->
<!--   pch = 12, -->
<!--   xlab = "Fitted values", -->
<!--   ylab = "Standardised residuals" -->
<!-- ) -->
<!-- ``` -->

<!-- ## QQ Plot -->

<!-- ```{r} -->
<!-- qqnorm(resid(mod), pch = 16) -->
<!-- qqline(resid(mod)) -->
<!-- ``` -->

<!-- ## Leverage plot -->

<!-- ```{r} -->
<!-- #Calculate leverage -->
<!-- lev <- hat(model.matrix(mod)) -->
<!-- #Plot leverage against standardized residuals -->
<!-- plot( -->
<!--   resid(mod, type = "pearson") ~ lev, -->
<!--   las = 1, -->
<!--   ylab = "Standardized residuals", -->
<!--   xlab = "Leverage" -->
<!-- ) -->
<!-- ``` -->
<!-- ::: -->

<!-- ```{r} -->
<!-- summary(mod) -->
<!-- ``` -->


<!-- As we can see from `summary(mod)` there is no significant difference in islet area of long-duration patients with respect to non-diabetic patients when accounting for correlation on the patient and image level. -->
